---
title: "Exercise 7"
output: html_document
date: "2025-02-26"
---
For this exercise, you will practice detecting if there are significant differences in a continuous variable across two or more levels of catgorical (nominal and ordinal) variables. These kinds of statistical procedures are generally referred to as bivariate statistics.

1. If you need to reload the NHANES data and label the ordinal and nominal variables as factor variables, you should run the code chunk below:
```{r}
#Use this code chunk to download the NHANES 2017-2018 database and to label the nominal and ordinal variables as factor variables (if necessary):
NHANESdata <- read.csv("https://raw.githubusercontent.com/malenamalka/ANTH241-SP25/refs/heads/main/data/NHANES2017-28")
factor(c(NHANESdata$Gender, NHANESdata$Race, NHANESdata$HealthIns))
factor(c(NHANESdata$Education,NHANESdata$HHEducation,NHANESdata$HHIncome,NHANESdata$FeelBad,NHANESdata$HHFoodSec),order=TRUE)
factor(NHANESdata$BMIcat, order =TRUE, levels = c("Underweight","Healthy","Overweight","Obese"))
head(NHANESdata)
```

2. Run the code chunk below to install and load the packages that we will use in this exercise:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
#install.packages("dunn.test")
library(dunn.test)
```
Test whether age varies by gender in the NHANES sample. What is the null hypothesis? What is the alternative hypothesis? 
null hypothesis is that age doesn't vary by gender
alternative hypothesis is that there is some significant variation by gender. 
```{r}
hist(NHANESdata$Age)
NHANESdata%>%
  drop_na(Age)%>%
  group_by(Gender)%>%
  summarise(mean(Age), sd(Age), count=n())
var.test(Age~Gender, data=NHANESdata, alternative="two.sided")
t.test(Age~Gender, var.equal=TRUE, data=NHANESdata)
```
The data refute the alternative hypothesis; p > 0.05

3. T-tests compare the variation of a continuous variable across two levels of a categorical (nominal and ordinal) variable. 
```{r}
#3a. What are the means, standard deviations, and counts for BMI by gender?
NHANESdata %>%
  drop_na(BMI)%>%
  group_by(Gender)%>%
  summarise(mean(BMI), sd(BMI), count=n())

#Let's hypothesize that there is a significant difference in BMI by gender in the NHANES 2017-2018 sample. What is the null hypothesis? What is the alternative hypothesis? We will test this hypothesis using the method shown in S&B, section 7.1.2
#1. histogram of continuous variable: hist(data$contvar)
hist(NHANESdata$BMI)
#2. F-test to see if the variance in the continuous variable is the same for the different levels of the categorical variable: 
#var.test(Contvar~Catvar, data=, alternative="two.sided")
var.test(BMI~Gender, data=NHANESdata, alternative="two.sided")
#3. According to S&B, if the variances are not equal, we should run the T-test anyways: t.test(Contvar~Catvar, var.equal=FALSE, data=)
t.test(BMI~Gender, var.equal=FALSE, data=NHANESdata)
```
Question: Does the above T-test support the null hypothesis or the alternative hypothesis?
The above T-test supports the alternative hypothesis as p<0.05
```{r}
#3b: Does BMI differ in people with and without health insurance? What is the null hypothesis? What is the alternative hypothesis?
#1. Find the descriptive statistics (mean, sd, and counts) of BMI by presence/absence of health insurance. 
#a. Below, I filter out levels of health insurance that are above 2 into a new database to ensure that the 7s and 9s are not part of the analysis.
NHANES_healthins <- NHANESdata%>%
  filter(HealthIns<3)
NHANES_healthins%>%
  drop_na(BMI)%>%
  group_by(HealthIns)%>%
  summarise(mean(BMI), sd(BMI), count=n())
#2. Run a test to see if the variances of the continuous variable are equal across different levels of the categorical variable:
#var.test(Contvar~Catvar, data=, alternative="two.sided")
var.test(BMI~HealthIns, data=NHANES_healthins, alternative="two.sided")
#3. We will run the T-test with the following code:
#t.test(Contvar~Catvar, var.equal=TRUE, data=)
t.test(BMI~HealthIns, var.equal=TRUE, data=NHANES_healthins)
```
Does the above T-test support or refute the alternative hypothesis?
The T-test supports the alternative hypothesis, p<0.05

3b. Sedentary and health insurance
```{r}
NHANES_sedentary1 <- NHANES_healthins%>%
  filter(Sedentary<1321)
NHANES_sedentary1%>%
  drop_na(Sedentary)%>%
  group_by(HealthIns)%>%
  summarise(mean=mean(Sedentary), sd=sd(Sedentary), count=n())
var.test(Sedentary~HealthIns, data=NHANES_sedentary1, alternative="two.sided")
t.test(Sedentary~HealthIns, var.equal=FALSE, data=NHANES_sedentary1, na.rm=TRUE)
```


3c. Generate a hypothesis that tests if daily caloric intake differs in people who identify as female from people who identify as male. What are the null and alternative hypotheses? Run a T-test by replicating the methods above. What are your results? Do your results support or refute the alternative hypothesis?
```{r}
NHANESdata%>%
  drop_na(DR1TKCAL)%>%
  group_by(Gender)%>%
  summarise(mean=mean(DR1TKCAL), sd=sd(DR1TKCAL), count=n())
var.test(DR1TKCAL~Gender, data=NHANESdata, alternative="two.sided")
t.test(DR1TKCAL~Gender, var.equal=TRUE, data=NHANESdata)
```
The null hypothesis for this test is that there is no significant difference in daily caloric intake across gender. The alternative hypothesis, on the other hand, states that there is some significant difference. The test done above provides evidence in favour of the alternative hypothesis as the p-value produced from the t-test < 0.05. We can thus say, with a 95% confidence interval, that there is a significant difference in daily caloric intake by gender. 

4. Testing for significant differences in a continuous variable across three or more categorical levels. Both S&B and Kabacoff introduce you to analysis of variance (ANOVA) as the statistical method to make these comparisons. What they show you is all good. However, I would like to introduce you to a non-parametric test that is more appropriate to the NHANES data as it applies to variables that are not normally distributed. This method is similar to ANOVA, it is easy to implement, and it provides results that are easy to interpret. 

The non-parametric statistical test that is a substitute for ANOVA is called the Kruskal-Wallis test. It shows if there is a significant difference in the means and variance of a continuous variable across three or more categorical levels when the continuous variable is not normally distributed. After the Kruskal-Wallis test, a post-hoc test called the Dunn Test, identifies the specific bivariate comparisons that significantly differ from each other.

Here is an example:
```{r}
#4a. Does daily caloric intake differ by BMI categories? What is the null hypothesis? What is the alternative hypothesis?
#1. Generate summary statistics of daily caloric intake by BMI categories.
NHANESdata%>%
  drop_na(DR1TKCAL)%>%
  group_by(BMIcat)%>%
  summarise(mean=mean(DR1TKCAL), sd=sd(DR1TKCAL), count=n())

#2. Generate a boxplot of daily caloric intake by BMI categories to visualize the summary statistics
ggplot(NHANESdata, aes(x=BMIcat, y=DR1TKCAL)) + geom_boxplot(fill="green") + labs(x="BMI category", y="calories consumed per day", title="daily calories by BMIcat")
#3. Run the Kruskal-Wallis test and the post-hoc Dunn test to see if there are any significant differences in daily caloric intake by BMI categories. The code for the Kruskal-Wallis and Dunn test is: 
#dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANESdata$DR1TKCAL, g=NHANESdata$BMIcat, kw=TRUE)
```
Does the above test support or refute the alternative hypothesis? If it supports the alternative hypothesis, which categorical levels significantly differ from each other? Explain your findings.
This test supports the alternative hypothesis in relation to those in the underweight category. However, for the other three categories, the null hypothesis was supported in relation to each other. 

```{r}
#4b. Does the ratio of monthly income to poverty differ by the head of household's education level? What is the null hypothesis? What is the alternative hypothesis?
#1. Generate descriptive statistics of RFITP by HHEducation levels.
NHANESdata%>%
  drop_na(RFITP)%>%
  group_by(HHEducation)%>%
  summarise(mean=mean(RFITP), sd=sd(RFITP), count=n())
#2. Run the Kruskal-Wallis and Dunn tests:
dunn.test(x=NHANESdata$RFITP, g=NHANESdata$HHEducation, kw=TRUE)
```
Explain your results here.
According to the results of the KW test conducted above, we can say that there is a statistically significant difference in ratio of income to poverty index by the level of education achieved by the head of household. We can say that there is significant difference across all categories involved, as the p-value for each of them is far less than 0.05. 

#5. Run your own Kruskal-Wallis and Dunn test below based on comparing a continuous variable across a nominal or ordinal variable with three or more levels. Show your work in the code chunk and interpret your findings.
In the tests below, we will be evaluating Sedentary activity (in minutes) by racial categories. 

```{r}
NHANES_sedentary <- NHANESdata%>%
  filter(Sedentary<1321)
NHANES_sedentary%>%
  drop_na(Sedentary)%>%
  group_by(Race)%>%
  summarise(mean=mean(Sedentary), sd=sd(Sedentary), count=n())
dunn.test(x=NHANES_sedentary$Sedentary, g=NHANES_sedentary$Race, kw=TRUE)
```
The results indicated above by the Kruskal-Wallis test demonstrate a variety of findings. Across the majority of results there is a statistically significant difference in minutes spend in sedentary activity by racial category. Those that do not demonstrate a significant difference primarily fall under the column of MR, or Mixed Race. There is no statisitcally significant difference between the sedentary activity levels of those who identify as Mixed Race and Non-Hispanic Asian; MR and Non-Hispanic Black; MR and Non-Hispanic White. Aside from those involving Mixed Race individuals, there is no significant difference between those who identify as Mexican-American and those who identify as Other Hispanic and between those who identify as Non-Hispanic Asian and Non-Hispanic Black individuals. Across all other categories, we can see that there is a statistically significant difference. 